---
title: "Ontology Models"
author: "Roberto Viviani, Karin Labek, University of Innsbruck"
date: "2024-04-18"
output:
  html_document:
    df_print: paged
---

This markdown contains the code used for the results of the manuscript "AI-driven assessment of semantic similarity and association of responses between the 'Big Five' and DSM-5/ICD-11 personality traits", by Karin Labek, Sandra Bennett-Long, and Roberto Viviani.


```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.dim = c(8,5.2))
library(ggplot2)
library(dplyr)
library(tidyr)

# utilities to decode embeddings
source("embedding_utilities.R") 
embeddings <- loadEmbeddings("embeddings_openAI_large_de.csv")

#when we read the scales, we drop the extranumerary NEO items
scales <- read.csv("scales_data_extended.csv") |>  
  select(-NEOFFI61:-NEOFFI66)
cormx <- cor(as.matrix(scales |> select(starts_with("NEOFFI"))), 
    as.matrix(scales |> select(starts_with("PID5BF"))))
cordata <- data.frame(corrs = as.vector(cormx), cosdist = as.vector(embeddings$cdist),
                      typeNEO = embeddings$NEO$type, 
                      typePID = rep(embeddings$PID$type, each = nrow(embeddings$NEO)),
                      itemNEO = embeddings$NEO$itemID, 
                      itemPID = rep(embeddings$PID$itemID, each = nrow(embeddings$NEO)))
# typos in the dataset
cordata[cordata$typePID == "eccentrictiy", "typePID"] <- "eccentricity" 
cordata[cordata$typePID == "intimicy avoidance", "typePID"] <- "intimacy avoidance"
cordata$traitPID <- traits[cordata$typePID]

# create a group ordering for PID facets that respects traits
library(forcats)
traits <- as.integer(as.factor(cordata$traitPID))
traits <- traits * -20 + as.integer(as.factor(cordata$typePID))
cordata$typePID <- fct_reorder(cordata$typePID, traits)

rm(cormx)

# plotting utilities
source("plot_utilities.R")

# stat utilities
source("stats_utilities.R")

pidtraits <- c("PIDAnankastia", "PIDAntagonism", "PIDDetachment",
               "PIDDisinhibition", "PIDNegativeAffect", "PIDPsychoticism")
traitnames <- c("Anankastia", "Antagonism", "Detachment",
               "Disinhibition", "NegativeAffect", "Psychoticism")
```

## Semantic analysis and prediction of association of responses between scales

### Figure 1

```{r, message = FALSE}
ggplot(
  cordata |> group_by(typeNEO, traitPID) |> summarize(corr = mean(corrs)),
  aes(y = corr, x = traitPID, fill = typeNEO)) + 
  geom_col(position = "dodge", alpha = 0.8, color = "white") +
  scale_fill_viridis_d() + 
  labs(fill = "NEO trait") + 
  xlab("PID domain") + ylab("avg correlation") + theme_grey() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
ggsave("PIDxNEO.pdf", width = 150, height = 95, units = "mm")

```


```{r, message = FALSE, warning = FALSE}
ggplot(
  cordata |> group_by(typeNEO, traitPID) |> summarize(corr = mean(corrs), dist = mean(cosdist)),
  aes(y = abs(corr), x = dist)) + 
  geom_point(aes(color = typeNEO, shape = traitPID), alpha = 0.8, size = 4) +
  geom_smooth(method = "glm", method.args = c("family"="quasibinomial"), 
    alpha = 0.2, color = "darkgrey") + 
  scale_color_viridis_d() + 
  labs(color = "NEO trait", shape = "PID domain") + 
  xlab("semantic distance") + ylab("avg correlation") + theme_classic()
ggsave("subscales_openAI_large.pdf", width = 150, height = 95, units = "mm")

summary(glm(abs(corr) ~ dist, 
            data = cordata |> group_by(typeNEO, traitPID) |> summarize(corr = mean(corrs), dist = mean(cosdist)), 
            family = quasibinomial))
```

```{r, message = FALSE}
ggplot(
  cordata |> group_by(typeNEO, traitPID) |> summarize(corr = mean(corrs), dist = mean(cosdist)),
  aes(y = corr, x = dist)) + 
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE) +
  geom_point(aes(color = typeNEO, shape = traitPID), alpha = 0.8, size = 4) +
  geom_smooth(aes(color = typeNEO), method = "lm", se = FALSE) + 
  scale_color_viridis_d() + 
  labs(color = "NEO trait", shape = "PID domain") + 
  xlab("semantic distance") + ylab("avg correlation") + theme_classic()
ggsave("subscales_signed_openAI_large.pdf", width = 150, height = 95, units = "mm")

```

## Figure 2, Table 1

Please see the separate markdown AnalysisComparison.

## Analysis of NEO traits, Figures 3-5

### Neuroticism

```{r neuro_model, message = FALSE }
corrs <- pearscorr(scales, "NEOneuro", pidtraits)
print(corrs)
p <- barplf(corrs, "Neuroticism"); p

ggsave("neuroticism_openAI_large_corrs.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")

```

```{r neuroticism, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Neuroticism")
p <- bxplf(plotdata); p
ggsave("neuroticism_openAI_large_boxes.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")
anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "NegativeAffect")
p <- addneoline(p, plotdata, "Detachment")
p <- addneoline(p, plotdata, "Disinhibition")
p <- addneoline(p, plotdata, "Antagonism")
p + geom_text(data = filter(plotdata, cosdist < 0.43 | corrs > 0.5), 
            aes(label = typePID), alpha = 0.4) + 
  labs(text = element_text(size = 16)) +
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

ggsave("neuroticism_openAI_large.pdf",  width = 140, height = 120, units = "mm")

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```

### Extraversion

```{r extraversion_model, message = FALSE }
corrs <- pearscorr(scales, "NEOextra", pidtraits)
print(corrs)
p <- barplf(corrs, "Extraversion"); p

ggsave("extraversion_openAI_large_corrs.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")

```

```{r extraversion, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Extraversion")
p <- bxplf(plotdata); p
ggsave("extraversion_openAI_large_boxes.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")
anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "Detachment")
p + geom_text(data = filter(plotdata, (cosdist < 0.5 & corrs < -0.2)), 
            aes(label = typePID), alpha = 0.4) + 
  labs(text = element_text(size = 16)) +
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

ggsave("extraversion_openAI_large.pdf", width = 140, height = 120, units = "mm")

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```


### Conscientiousness

```{r conscientiousness_model, message = FALSE }
corrs <- pearscorr(scales, "NEOgew", pidtraits)
print(corrs)
p <- barplf(corrs, "Conscientiousness"); p

ggsave("conscientiousness_openAI_large_corrs.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")

```


```{r conscientiousness, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Conscientiousness")
p <- bxplf(plotdata); p
ggsave("conscientiousness_openAI_large_boxes.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")
anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "Anankastia")
p <- addneoline(p, plotdata, "Disinhibition")
p + geom_text(data = filter(plotdata, cosdist < 0.45), 
            aes(label = typePID), alpha = 0.4) +
    geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

ggsave("conscientiousness_openAI_large.pdf", width = 140, height = 120, units = "mm")

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```

```{r conscientiousness_items, message = FALSE}
p + geom_text(data = filter(plotdata, cosdist < 0.45), 
            aes(label = paste(itemNEO, itemPID, sep=":")), alpha = 0.3)
```

```{r consc_detail_openAI, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Conscientiousness", traitPID == "Anankastia")
p <- baseplot(plotdata, TRUE)
p <- addneoline(p, plotdata, "Anankastia")
p + geom_text(data = filter(plotdata, cosdist < 0.45), 
            aes(label = typePID), alpha = 0.4) + 
  scale_color_viridis_d(name = "PID facet", end = 0.9) +
  labs(title = "Conscientiousness in Anankastia", colour = "PID facet") +
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE) +
  theme(legend.position = "right")

ggsave("conscientiousness_detail_openAI_large.pdf", width = 100, height = 60, units = "mm")

```


```{r consc_detail_gecko_en, message = FALSE}
openAIemb <- loadEmbeddings("embeddings_OpenAI_large_en.csv")
cordataAIen <- cordata
cordataAIen$cosdist <- as.vector(openAIemb$cdist)
plotdata <- filter(cordataAIen, typeNEO == "Conscientiousness", traitPID == "Anankastia")
p <- baseplot(plotdata, TRUE)
p <- addneoline(p, plotdata, "Anankastia")
p + geom_text(data = filter(plotdata, cosdist < 0.45 | corrs > 0.4), 
            aes(label = typePID), alpha = 0.4) + 
  scale_color_viridis_d(name = "PID facet", end = 0.9) +
  labs(title = "Conscientiousness in Anankastia", colour = "PID facet") + 
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE) +
  theme(legend.position = "right")
rm("cordataAIen", "openAIemb")
ggsave("conscientiousness_detail_openAI_large_en.pdf", width = 100, height = 60, units = "mm")

```

### Agreeableness

```{r agreeableness_model, message = FALSE }
corrs <- pearscorr(scales, "NEOvertr", pidtraits)
print(corrs)
p <- barplf(corrs, "Agreeableness"); p

ggsave("agreeableness_openAI_large_corrs.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")

```


```{r agreeableness, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Agreeableness")
p <- bxplf(plotdata); p
ggsave("agreeableness_openAI_large_boxes.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")
anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "Antagonism")
p <- addneoline(p, plotdata, "Psychoticism")
p <- addneoline(p, plotdata, "Detachment")
p + geom_text(data = filter(plotdata, cosdist < 0.5, corrs < -0.2), 
            aes(label = typePID), alpha = 0.4) +
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

ggsave("agreeableness_openAI_large.pdf", width = 140, height = 120, units = "mm")

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```


### Openness to experience

```{r openness_model, message = FALSE }
corrs <- pearscorr(scales, "NEOoff", pidtraits)
print(corrs)
p <- barplf(corrs, "Openness"); p

ggsave("openness_openAI_large_corrs.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")

```

```{r openness, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Openness")
p <- bxplf(plotdata); p
ggsave("openness_openAI_large_boxes.pdf", 
       plot = p + labs(title = NULL) + theme(legend.position = "null", 
                        text = element_text(size = 14)), 
       width = 45, height = 60, units = "mm")
anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "Detachment")
p <- addneoline(p, plotdata, "Psychoticism")
p + geom_text(data = filter(plotdata, corrs < -0.2), 
            aes(label = typePID), alpha = 0.4) +
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

ggsave("openness_openAI_large.pdf", width = 140, height = 120, units = "mm")

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```

