---
title: "Ontology Models"
author: "Roberto"
date: "2024-04-18"
output: word_document
---

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Rcpp)
# define matlab's pdist2
cppFunction('NumericMatrix crossdist(NumericMatrix m1, NumericMatrix m2) {

  int nrow1 = m1.nrow();
  int nrow2 = m2.nrow();
  int ncol = m1.ncol();
 
  if (ncol != m2.ncol()) {
    throw std::runtime_error("Incompatible number of dimensions");
  }
 
  NumericMatrix out(nrow1, nrow2);
 
  for (int r1 = 0; r1 < nrow1; r1++) {
    for (int r2 = 0; r2 < nrow2; r2++) {
      double total = 0;
      for (int c12 = 0; c12 < ncol; c12++) {
        total += pow(m1(r1, c12) - m2(r2, c12), 2);
      }
      out(r1, r2) = sqrt(total);
    }
  }
 
  return out;

}')

embmx <- function(frm) {
  mx = matrix(nrow = nrow(frm), ncol = 3072)
  for (count in 1 : nrow(frm)) {
    eb <- frm[count,"embedding"]
    em <- sub("\\[", "c\\(", eb)
    em <- sub("\\]", "\\)", em)
    c <- eval(str2expression(em))
    mx[count,] <- c
  }
  mx
}

embeddings <- read.csv("scales_openAI_large.csv")

library(dplyr)
NEO <- filter(embeddings, scaleID == "NEO") |> select(itemID, type, embedding)
PID <- filter(embeddings, scaleID == "PID") |> select(itemID, type, embedding)
NEOmx <- embmx(NEO)
PIDmx <- embmx(PID)
rm(embeddings)

scales <- read.csv("scales_data.csv")
cormx <- cor(as.matrix(scales |> select(starts_with("NEO"))), 
    as.matrix(scales |> select(starts_with("PID"))))

```

We plot correlation on cosine distance of emebddings, to see if distance predicts correlation. We consider different NEO subscales separately.

```{r, message = FALSE }
cdist <- crossdist(NEOmx, PIDmx)
cormx <- cor(as.matrix(scales |> select(starts_with("NEO"))), 
    as.matrix(scales |> select(starts_with("PID"))))
cordata <- data.frame(corrs = as.vector(cormx), cosdist = as.vector(cdist),
                      itemNEO = NEO$type, itemPID = rep(PID$type, each = nrow(NEO)))

library(ggplot2)
ggplot(cordata, aes(cosdist, corrs, color = itemNEO)) + geom_point(shape=21) + 
  geom_smooth(aes(group = itemNEO), method = "lm", se = F) + 
  theme_classic()

```

This is as expected, except for conscientiousness. Here, we can show that with lower semantic distance the associations diverge into two groups.

```{r, message = FALSE}
ggplot(filter(cordata, itemNEO == "Gewissenhaftigkeit"), 
       aes(cosdist, corrs)) + 
  geom_point(aes(fill = itemPID, color = itemPID), alpha = 0.5, shape=21) + 
  geom_smooth(aes(group = corrs > 0), color = "grey", method = "lm", se = F) + 
  theme_classic()

```

Furthermore, the items that associated positively with PID subscales concern rigidity, separation anxiety, orderliness. Distractability and manipulativenes/irresponsibility are negatively associated with conscientiousness, meaning individuals who are more conscienscious are less distractable and irresponsible.
