---
title: "Supplementary results C (English embeddings)"
author: "Roberto Viviani, Karin Labek, University of Innsbruck"
date: "2024-04-18"
output: pdf_document
---

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.dim = c(8,5.2))
library(ggplot2)
library(dplyr)
library(tidyr)

# utilities to decode embeddings
source("embedding_utilities.R") 
embeddings <- loadEmbeddings("embeddings_openAI_large_en.csv")

#when we read the scales, we drop the extranumerary NEO items
scales <- read.csv("scales_data_extended.csv") |>  
  select(-NEOFFI61:-NEOFFI66)
cormx <- cor(as.matrix(scales |> select(starts_with("NEOFFI"))), 
    as.matrix(scales |> select(starts_with("PID5BF"))))
cordata <- data.frame(corrs = as.vector(cormx), cosdist = as.vector(embeddings$cdist),
                      typeNEO = embeddings$NEO$type, 
                      typePID = rep(embeddings$PID$type, each = nrow(embeddings$NEO)),
                      itemNEO = embeddings$NEO$itemID, 
                      itemPID = rep(embeddings$PID$itemID, each = nrow(embeddings$NEO)))
# typos in the dataset
cordata[cordata$typePID == "eccentrictiy", "typePID"] <- "eccentricity" 
cordata[cordata$typePID == "intimicy avoidance", "typePID"] <- "intimacy avoidance"
cordata$traitPID <- traits[cordata$typePID]

# create a group ordering for PID facets that respects traits
library(forcats)
traits <- as.integer(as.factor(cordata$traitPID))
traits <- traits * -20 + as.integer(as.factor(cordata$typePID))
cordata$typePID <- fct_reorder(cordata$typePID, traits)

rm(cormx)

# plotting utilities
source("plot_utilities.R")

# stat utilities
source("stats_utilities.R")

pidtraits <- c("PIDAnankastia", "PIDAntagonism", "PIDDetachment",
               "PIDDisinhibition", "PIDNegativeAffect", "PIDPsychoticism")
traitnames <- c("Anankastia", "Antagonism", "Detachment",
               "Disinhibition", "NegativeAffect", "Psychoticism")
```

This document contains analyses that supplement the main text of "AI-driven assessment of semantic similarity and association of responses between the 'Big Five' and DSM-5/ICD-11 personality traits", by Karin Labek, Sandra Bennett-Long, and Roberto Viviani. This document repeated the analysis in the English language. We conducted these analyses to verify that the English version of the embedding performed similarly to the German version. The detection of a divergence of estimates of semantic similarity for items in the consciousness NEO trait was reported in the main text.

This is not a replication of the results with a new English-speaking sample. The sample is the same as in the main analysis.

```{r, message = FALSE, warning = FALSE}
ggplot(
  cordata |> group_by(typeNEO, traitPID) |> summarize(corr = mean(corrs), dist = mean(cosdist)),
  aes(y = abs(corr), x = dist)) + 
  geom_point(aes(color = typeNEO, shape = traitPID), alpha = 0.8, size = 4) +
  geom_smooth(method = "glm", method.args = c("family"="quasibinomial"), 
    alpha = 0.2, color = "darkgrey") + 
  scale_color_viridis_d() + 
  labs(color = "NEO trait", shape = "PID domain") + 
  xlab("semantic distance") + ylab("avg correlation") + theme_classic()

summary(glm(abs(corr) ~ dist, 
            data = cordata |> group_by(typeNEO, traitPID) |> summarize(corr = mean(corrs), dist = mean(cosdist)), 
            family = quasibinomial))
```

One can see that the model chosen in the analysis in the main text performed similarly to the English model, which attained only a slightly superior test statistics (t = -5.5, p < 0.001). Also the association between subscales is similar:

```{r, message = FALSE}
ggplot(
  cordata |> group_by(typeNEO, traitPID) |> summarize(corr = mean(corrs), dist = mean(cosdist)),
  aes(y = corr, x = dist)) + 
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE) +
  geom_point(aes(color = typeNEO, shape = traitPID), alpha = 0.8, size = 4) +
  geom_smooth(aes(color = typeNEO), method = "lm", se = FALSE) + 
  scale_color_viridis_d() + 
  labs(color = "NEO trait", shape = "PID domain") + 
  xlab("semantic distance") + ylab("avg correlation") + theme_classic()

```


### Neuroticism

```{r neuroticism, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Neuroticism")
p <- bxplf(plotdata); p

anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "NegativeAffect")
p <- addneoline(p, plotdata, "Detachment")
p <- addneoline(p, plotdata, "Disinhibition")
p <- addneoline(p, plotdata, "Antagonism")
p + geom_text(data = filter(plotdata, cosdist < 0.43 | corrs > 0.5), 
            aes(label = typePID), alpha = 0.4) + 
  labs(text = element_text(size = 16)) +
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```
### Extraversion

Extraversion appears to be characterized by the selective negative association of some items with anhedonia, anxiety, and withdrawal. We see the lower part of the plot, containing negative associations, being occupied by these items. 

```{r extraversion, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Extraversion")
p <- bxplf(plotdata); p

anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "Detachment")
p + geom_text(data = filter(plotdata, (cosdist < 0.5 & corrs < -0.2)), 
            aes(label = typePID), alpha = 0.4) + 
  labs(text = element_text(size = 16)) +
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```

### Conscientiousness

As already mentioned in the main text, the English version of the conscientiousness gives a different result of semantic similarity with the anankastia PID domain.

```{r conscientiousness, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Conscientiousness")
p <- bxplf(plotdata); p

anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "Anankastia")
p <- addneoline(p, plotdata, "Disinhibition")
p + geom_text(data = filter(plotdata, cosdist < 0.45), 
            aes(label = typePID), alpha = 0.4) +
    geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```

### Agreeableness

```{r agreeableness, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Agreeableness")
p <- bxplf(plotdata); p

anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "Antagonism")
p <- addneoline(p, plotdata, "Psychoticism")
p <- addneoline(p, plotdata, "Detachment")
p + geom_text(data = filter(plotdata, cosdist < 0.5, corrs < -0.2), 
            aes(label = typePID), alpha = 0.4) +
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```


### Openness to experience

```{r openness, message = FALSE}
plotdata <- filter(cordata, typeNEO == "Openness")
p <- bxplf(plotdata); p

anovadist(plotdata)

p <- baseplot(plotdata)
p <- addneoline(p, plotdata, "Detachment")
p <- addneoline(p, plotdata, "Psychoticism")
p + geom_text(data = filter(plotdata, corrs < -0.2), 
            aes(label = typePID), alpha = 0.4) +
  geom_hline(yintercept = 0, color = "darkgrey", show.legend = FALSE)

pearscorr(plotdata |> select(traitPID, corrs, cosdist) |> pivot_wider(names_from = traitPID, values_from = cosdist), "corrs", traitnames)

```
